---
title: "Spatial Demography with Open Source Tools"
author: "Kyle Walker"
date: 'May 5, 2021'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE, tigris_use_cache = TRUE)
options(cancensus.api_key = Sys.getenv("CANCENSUS_API_KEY"))
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_xaringan(
  title_slide_background_color = "white",
  title_slide_text_color = "#4d1979",
  text_color = "black",
  header_color = "#4d1979",
  inverse_background_color = "#4d1979",
  text_font_family = "Gotham Narrow",
  text_font_size = "1.2rem",
  header_font_family = "Helvetica",
  header_font_weight = "bold",
  link_color = "#4d1979",
  code_inline_color = "#4d1979",
  title_slide_background_image = "img/TCULogo_purple_5X7-01.jpg",
  title_slide_background_size = "250px",
  title_slide_background_position = "95% 95%"
)
```

## Software and spatial demography

<img src=img/commercial.png style="width: 700px">

---

## Problems with traditional workflows

* Point-and-click software can inhibit reproducibility and collaboration

--

* Even if an analysis is scripted: 

--

  - The use of commercial software makes workflows inaccessible to analysts who cannot afford the license fees;
  
--
  
  - Data access is rarely automated and is difficult to reproduce; 
  
--
  
  - Data wrangling, modeling, and visualization are "siloed" in different commercial software packages

---

## Open source workflows

* __Data sources__: open data APIs

* __Data models__: GeoJSON, simple features (attributes + geometry in single file/object)

* __Modeling & analytics tools__: sf + spdep (R), GeoPandas + PySAL (Python)

* __Presentation tools__: cartography (ggplot2, tmap) + dashboarding (Shiny) in same software environment as analysis

---

## Open data APIs

.pull-left[

* Open data APIs allow analysts to programmatically access data resources

* Examples include official government APIs (US Census) and third-party APIs (e.g. Canada's CensusMapper)

* Software tools like {censusapi} and {tigris} (US) and {cancensus} (Canada) allow for programmatic access to data resources

]

.pull-right[

```{r, results = 'hide'}
library(cancensus)

vancouver_extract <- get_census(
  dataset = "CA16",
  regions = list(CSD = "5915022"),
  vectors = c("v_CA16_1364", 
              "v_CA16_2397",
              "v_CA16_244"),
  level = "CT",
  geo_format = "sf",
  labels = "short"
)
```

]

---


## Data models

.pull-left[

* Frameworks like the __tidyverse__ (R) and __pandas__ (Python) are widely used for data representation in open-source workflows

* Geospatial extensions to these frameworks (__sf__ for R, __GeoPandas__ for Python) allow spatial demographers to analyze data in similar ways to a regular dataset without dedicated GIS software

]

.pull-right[

```{r}
library(tidyverse)

vancouver_data <- vancouver_extract %>%
  transmute(
    tract_id = GeoUID,
    pct_english = 100 * (v_CA16_1364 / Population),
    median_income = v_CA16_2397,
    pct_65_up = 100 * (v_CA16_244 / Population)
  )

```
]

---

```{r}
vancouver_data
```

---

```{r}
plot(vancouver_data$geometry)
```

---

## Modeling & analytics

.pull-left[

* Models fit to spatial data require modeling _spatial dependence_, often represented with a neighborhood spatial weights matrix

* Example workflow: regionalization with the SKATER algorithm, equivalent to ArcGIS's Spatially Constrained Multivariate Clustering

]


.pull-right[

```{r, echo = FALSE}
library(spdep)
set.seed(123456)

nb <- poly2nb(vancouver_data,
              queen = FALSE)

van_coords <- vancouver_data %>%
  st_centroid() %>%
  st_coordinates()

plot(vancouver_data$geometry)
plot(nb, 
     coords = van_coords, 
     add = TRUE, 
     col = "blue", 
     points = FALSE)
```

]

---

## Modeling and analytics

<iframe src="img/mapview.html" frameborder="0" seamless scrolling="no" height="450" width="800"></iframe>

---

## Presentation tools

<iframe src="img/ggiraph.html" frameborder="0" seamless scrolling="no" height="450" width="800"></iframe>

---

## Sharing and reproducibility

.pull-left[

* This entire workflow is fully reproducible - [just take a look at the source code](https://github.com/walkerke/paa2021/blob/main/code/paa_code.R) and run it for yourselves!

* Adoption of open science practices in spatial demography facilitates _collaboration_, _education_, and _innovation_

]

.pull-right[

<img src=img/github.png style="width: 400px">

]

---
class: middle, center, inverse

## Thank you!
